---
title: "Extending lumberjack"
author: "Mark van der Loo"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Adding your own loggers}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

There are many ways one may want to register changes in data.
That is why `lumberjack` allows you to write your own loggers. Here you can
read how to do it.

## The lumberjack logging API

In short, a logger is a _reference object_ with the following methods:

1. `$add(meta, input, output)`. This is a function that computes the
difference between `input` and `output` and adds it to a log. The `meta`
argument is a `list` with two elements:
    - `expr` The expression used to turn `input` into `output`
    - `src` The same expression, but turned into a string.
2. `$dump()` This function writes dumps the current logging info somewhere.
Often this will be a file, but it really can be any place where R can send data.

There are several systems in R to build such an object. We recommend
using [R6](https://cran.r-project.org/package=R6) classes or 
[Reference classes](http://adv-r.had.co.nz/R5.html).

Below we give an example in each system. The example loggers only register 
whether something has ever changed. A `dump` results in a simple message on
screen.


### R6 classes
An introduction to R6 classes can be found [here](https://cran.r-project.org/package=R6/vignettes/Introduction.html)

Let us define the 'trivial' logger.
```{r}
library(R6)
trivial <- R6Class("trivial",
  public = list(
    changed = NULL
  , initialize = function(){
      self$changed <- FALSE
  }
  , add = function(meta, input, output){
    self$changed <- self$changed | !identical(input, output)
  }
  , dump = function(){
    msg <- if(self$changed) "" else "not "
    cat(sprintf("The data has %schanged\n",msg))
  }
  )
)
```
Here is how to use it.
```{r}
library(lumberjack)
out <- women %L>% 
  start_log(trivial$new()) %L>%
  identity() %L>%
  dump_log(stop=TRUE)


out <- women %L>%
  start_log(trivial$new()) %L>%
  head() %L>%
  dump_log(stop=TRUE)
```

### Reference classes
Reference classes (RC) come with the R recommended `methods` package. 
An introduction can be found [here](http://adv-r.had.co.nz/R5.html). Here is how
to define the `trivial` logger as a reference class.
```{r}
library(methods)
trivial <- setRefClass("trivial",
  fields = list(
    changed = "logical"
  ),
  methods = list(
    initialize = function(){
      .self$changed = FALSE
    }
    , add = function(meta, input, output){
      .self$changed <- .self$changed | !identical(input,output)
    }
    , dump = function(){
      msg <- if( .self$changed ) "" else "not "
      cat(sprintf("The data has %schanged\n",msg))
    }
  )
)
```
And here is how to use it.
```{r}
library(lumberjack)
out <- women %L>% 
  start_log(trivial()) %L>%
  identity() %L>%
  dump_log(stop=TRUE)


out <- women %L>%
  start_log(trivial()) %L>%
  head() %L>%
  dump_log(stop=TRUE)

```

Observe that there are subtle differences between R6 and Reference classes (RC).

- In R6 the object is referred to with `self`, in RC this is done with `.self`.
- An R6 object is initialized with `classname$new()`, an RC object is initialized with `classname()`. 

## Advice for package authors

If you have a package that has interesting functionality that can be offered
also inside a logger, you might consider exporting a logger object that works
with `lumberjack`. To keep things uniform, we give the following advice.

### Documenting logging objects.
Most package authors use [roxygen2](https://cran.r-project.org/package=roxygen2) 
to generate documentation. Below is an example of how to document the class
and its methods. To show how to document arguments, we included and
`allcaps` argument in the dump function.

```
#' The trivial logger.
#' 
#' The trivial logger only registers whether something has changed at all.
#' A `dump` leads to an informative message on the console.
#'
#' @section Creating a logger:
#' \code{trivial$new()}
#' 
#' @section Dump options:
#' \code{$dump(allcaps)}
#' \tabular{ll}{
#'   \code{allcaps}\tab \code{[logical]} print message in capitals?
#' }
#' 
#' 
#' @docType class
#' @format An \code{R6} class object.
#' 
#' @examples
#' out <- women %L>%
#'  start_log(trivial$new()) %L>%
#'  head() %L>%
#'  dump_log(stop=TRUE)
#' 
#'
#' @export
trivial <- R6Class("trivial",
  public = list(
    changed = NULL
  , initialize = function(){
      self$changed <- FALSE
  }
  , add = function(meta, input, output){
    self$changed <- self$changed | !identical(input, output)
  }
  , dump = function(allcaps=FALSE){
    msg <- if(self$changed) "" else "not "
    msg <- sprintf("The data has %schanged\n",msg)
    if (allcaps) msg <- toupper(msg)
    cat(msg)
  )
)
```


### Adding lumberjack to the DESCRIPTION of your package

Once you have exported a logger, it is a good idea to add the line
```
Enhances: lumberjack
```
To your `DESCRIPTION` file. It can then be found by other users via lumberjack's
CRAN webpage.

## API overview

We briefly summarize the lumberjack API.

If you implement a logging object, the following methods are **mandatory**:

- `add(meta, input, output)`
    - `meta`: a list with components `expr` (an expression) and `src` (a string)
    - `input`: data before processing
    - `output`: data after processing
- `dump(...)` 
    - `...`: options to pass to other methods.

The following methods are **optional**

- `stop()` called by `stop_log()` before removing the logger from the data.








