---
title: "Getting started with lumberjack"
author: "Mark van der Loo"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc : true
vignette: >
  %\VignetteIndexEntry{Getting started with lumberjack}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
library(lumberjack)
```

This package allows you to monitor changes in your data as they get processed. For
this purpose it defines a few functions to start and stop logging, and the `lumberjack`:
an operator that logs.


### Installation

The basic version can be installed with
```{r,eval=FALSE}
install.packages("lumberjack")
```
This gives you basic logging functionality and is sufficient to apply your home-brewn loggers. If you want to use more advanced loggers that come with the
package, use
```{r, eval=FALSE}
install.packages("lumberjack", dependencies = TRUE)
```
This also installs the suggested packages and their dependencies when needed.

### The lumberjack workflow
So you want to know who does what to your data as it flows through your
process. Here's the workflow that allows you to do it using the `lumberjack`
package. 
```{r}
out <- women %>>%
  start_log() %>>%
  identity() %>>%
  head() %>>%
  dump_log(stop=TRUE)
read.csv("simple_log.csv")
```
Lets go through this step by step to see what happened. The start of the script 
defines an output varaiable `out` and passes `women` to the lumberjack (`%>>%`).
Next, the function `start_log` makes sure that logging starts from there. We are
now ready to start performing logged transformations on our dataset. First, we 
apply the `identity` function, which does exactly nothing. Then, the `head` 
function selects the first six rows in of the dataset and `dump_log(stop=TRUE)`
writes the log to a csv file, which we then read in. The optional argument 
`stop=TRUE` signals that no logging is necessary for following activities.

The resulting data frame contains a step number, a timestamp, the expression 
evaluated to transform the data, and an indicator wheter the data had changed at
all. As expected, the `identity` function hasn't changed anything and the
`head` function cuts of all records below the sixth row.


You have now seen the most important functions of the package. Let's summarize them.

- `start_log(data, log)`: start logging using possibly a custom logger (see next section)
- `%>>%`: the lumberjack. a logging-aware function composition operator ('pipe').
- `dump_log(data, stop, ...)`: dump the log for `data` (if present)
- `stop_log(data)`: stop logging.

All these functions are data-in, data-out. You are probably used to this from 
using [dplyr](http://dplyr.tidyverse.org/) or one of its siblings. However, the 
`lumberjack` functions are not limited to `data.frame`-like objects. In
principle, changes to any object type can be logged, but it depends on the
logger whether that will actually work.


### Changing the way things are logged

The `start_log` function adds a logging object as an attribute to its input data.
By default, this is the `simple` logger, which only checks whether data has
changed at all.




### Properties of the lumberjack


### A comparison with other pipe operators



