---
title: "Getting started with lumberjack"
author: "Mark van der Loo"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc : true
vignette: >
  %\VignetteIndexEntry{Getting started with lumberjack}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, echo=FALSE}
library(lumberjack)
```

This package allows you to monitor changes in data as they get processed. It
implements an easy-to-use and extensible logging framework, and comes with a few
data loggers implemented.

This vignette will show you how to get started and what the default loggers do. The [extending lumberjack](extending.html) vignette explains how to build your own loggers.



## Installation

The basic version can be installed with
```{r,eval=FALSE}
install.packages("lumberjack")
```

## The lumberjack workflow
So you want to know who does what to your data as it flows through your
process. Here's the workflow that allows you to do it using the `lumberjack`
package. 
```{r}
out <- women %>>%
  start_log() %>>%
  identity() %>>%
  head() %>>%
  dump_log(stop=TRUE)
read.csv("simple_log.csv")
```
Lets go through this step by step to see what happened. The start of the script 
defines an output varaiable `out` and passes `women` to the lumberjack (`%>>%`).
Next, the function `start_log` makes sure that logging starts from there. We are
now ready to start performing logged transformations on our dataset. First, we 
apply the `identity` function, which does exactly nothing. Then, the `head` 
function selects the first six rows in of the dataset and `dump_log(stop=TRUE)`
writes the log to a csv file, which we then read in. The optional argument 
`stop=TRUE` signals that no logging is necessary for following activities.

The resulting data frame contains a step number, a timestamp, the expression 
evaluated to transform the data, and an indicator wheter the data had changed at
all. As expected, the `identity` function hasn't changed anything and the
`head` function cuts of all records below the sixth row.


You have now seen the most important functions of the package. Let's summarize them.

- `start_log(data, log)`: start logging using possibly a custom logger (see next section)
- `%>>%`: the lumberjack. a logging-aware function composition operator ('pipe').
- `dump_log(data, stop, ...)`: dump the log for `data` (if present)
- `stop_log(data)`: stop logging.

All these functions are data-in, data-out. You are probably used to this from 
using [dplyr](http://dplyr.tidyverse.org/) or one of its siblings. However, the 
`lumberjack` functions are not limited to `data.frame`-like objects. In
principle, changes to any object type can be logged, but it depends on the
logger whether that will actually work -- most will expect a particular data
structure.

## Changing the logger

This is easy. For example, to use the builtin \code{cellwise} logger,
we just tell \code{start_log()} what logger to use. For this logger,
it is necessary to have a key column that identifies the rows uniquely.
```{r}
logfile <- tempfile(fileext = ".csv")
women$a_key <- seq_len(nrow(women))
out <- women %>>%
  start_log( log = cellwise$new(key="a_key") ) %>>%
  {.$height[1] <- 60; .} %>>%
  head(13) %>>%
  dump_log(file=logfile, stop=TRUE)

read.csv(logfile)
```




## Changing logger behaviour
There are two ways to change how a logge behaves. By setting options
at initialization and by setting options when dumping a log.


### Setting options for the logger

The `start_log` function adds a logging object as an attribute to its input 
data. By default, this is the `simple` logger, which only checks whether data 
has changed at all. The behaviour of this logger can be changed by 
passing options when it is created. To see this, have a look at the complete
call, as it is executed by default.
```{r}
dat <- start_log(women, log = simple$new())
```
The expression `simple$new()` creates a new logging object, and `start_log`
makes sure it is attached as an attribute to the copy of the `women` dataset
stored in `dat`. The simple logger has one option called `verbose`, that can be set when calling `$new`. The default is `TRUE`, here we set it to `FALSE`.
```{r}
dat <- start_log(women, log=simple$new(verbose=FALSE))
```
The effect is that no message is printed when the log is dumped to file.
```{r}
out <- dat %>>% identity() %>>% dump_log()
read.csv("simple_log.csv")
```

Note that the available options depend logger you use.

### Setting options for the output

For the simple logger, the default output file is `simple_log.csv` This can be
changed when calling `dump_log`. 
```{r}
out <- dat %>>% identity() %>>% dump_log(file="log_all_day.csv")
read.csv("log_all_day.csv")
```
In fact, `dump_log`  accepts any option that is accepted by `write.csv`, (except
`row.names`).

### Options for other loggers

Loggers can come in different forms. In principle, authors are free to
use R6 classes (as is done here), Reference classes, or anything else that
follows the lumberjack API. This means that the way tha logging objects are
initialized may vary from logger to logger. Check the documentation of a logger
to see how to operate it. The loggers included with this package are can be
found with `help('cellwise')` and `help('simple')`.

## Properties of the lumberjack

There are several function composition ('pipe') operators in the R community, 
including [magrittr](https://cran.r-project.org/package=magrittr), 
[pipeR](https://cran.r-project.org/package=pipeR) and 
[yapo](https://github.com/piccolbo/yapo). All have different behavior. 


The  lumberjack operator behaves as a simplified version of the `magrittr` pipe
operator. Here are some examples.
```r
# pass arguments to a function
1:3 %>>% mean()

# pass arguments using "."
TRUE %>>% mean(c(1,NA,3), na.rm = .)

# pass arguments to an expression, using "."
1:3 %>>% { 3 * .}

# in a more complicated expression, return "." explicitly
women %>>% { .$height <- 2*.$height; . }
```

The main differences are that 
- there is no assignment-pipe like `%<>%`.
- it does not allow you to define functions in the magrittr style: `a <- . %>% sin(.)`





